Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2024-01-08 article.tex
Remove empty subcaptions
Fix minipage widths
Fix text error
Move figure code to external files

\begin{figure}

\begin{minipage}[t]{\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{from}\NormalTok{ poincare }\ImportTok{import}\NormalTok{ Derivative, System, Variable, initial}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}

{\centering

\begin{minipage}[c]{0.33\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class}\NormalTok{ Decay(System):}
\NormalTok{    x: Variable }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
\NormalTok{    eq }\OperatorTok{=}\NormalTok{ x.derive() }\OperatorTok{\textless{}\textless{}} \OperatorTok{{-}}\NormalTok{x}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
%
\begin{minipage}[c]{0.33\linewidth}

{\centering

\[
\frac{dx}{dt} = -x
\]

}

\end{minipage}%
%
\begin{minipage}[c]{0.33\linewidth}

{\centering

\[
x(0) = 1
\]

}

\end{minipage}%

\subcaption{First-order system of an exponential decay. The variable
\texttt{x} stores the initial condition for \(x\), and the variable
\texttt{eq} stores the rate equation for \(x\).}

}

\subcaption{\label{}}
\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}

{\centering

\begin{minipage}[c]{0.33\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class}\NormalTok{ Oscillator(System):}
\NormalTok{    x: Variable }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
\NormalTok{    v: Derivative }\OperatorTok{=}\NormalTok{ x.derive(initial}\OperatorTok{=}\DecValTok{0}\NormalTok{)}
\NormalTok{    eq }\OperatorTok{=}\NormalTok{ v.derive() }\OperatorTok{\textless{}\textless{}} \OperatorTok{{-}}\NormalTok{x}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
%
\begin{minipage}[c]{0.33\linewidth}

{\centering

\[
\frac{d^2x}{dt^2} = -x
\]

}

\end{minipage}%
%
\begin{minipage}[c]{0.33\linewidth}

{\centering

\[
\begin{cases}
    x(0) &= 1 \\
    \frac{dx}{dt}(0) &= 0
\end{cases}
\]

}

\end{minipage}%

\subcaption{Second-order system of an harmonic oscillator, where the
variable \texttt{v} stores the derivative \(\frac{dx}{dt}\) and the rate
equation is specified for the derivative \texttt{v}.}

}

\subcaption{\label{}}
\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}

{\centering

\begin{minipage}[c]{0.33\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class}\NormalTok{ BigModel(System):}
\NormalTok{    x: Variable }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
\NormalTok{    linked }\OperatorTok{=}\NormalTok{ Decay(x}\OperatorTok{=}\NormalTok{x)}
\NormalTok{    independent }\OperatorTok{=}\NormalTok{ Decay(x}\OperatorTok{=}\DecValTok{2}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
%
\begin{minipage}[c]{0.33\linewidth}

{\centering

\[
\begin{cases}
    \frac{dx}{dt} = -x \\
    \frac{dy}{dt} = -y
\end{cases}
\]

}

\end{minipage}%
%
\begin{minipage}[c]{0.33\linewidth}

{\centering

\[
\begin{cases}
    x(0) &= 1 \\
    y(0) &= 2
\end{cases}
\]

}

\end{minipage}%

\subcaption{First-order system of two exponential decays by composition of
the \texttt{Decay} system. The subsystem \texttt{linked} has a reference
to the outer variable \texttt{x}, while the subsystem
\texttt{independent} defines a new variable with initial condition §2\$,
which on the corresponding mathematical expression was named \(y\).}

}

\subcaption{\label{}}
\end{minipage}%

\caption{\label{fig-poincare}Code and corresponding mathematical
expressions for different systems.}

\end{figure}

@@

\begin{figure}

\begin{minipage}[b]{0.50\linewidth}

{\centering

\includegraphics{src/ide/ide1.pdf}

}

\end{minipage}%
%
\begin{minipage}[b]{0.50\linewidth}

{\centering

\includegraphics{src/ide/ide2.pdf}

}

\end{minipage}%

\caption{\label{fig-ide}Screenshots of Visual Studio Code showing
tooltips and highlighted type errors. \texttt{Constant} are assigned
with \texttt{assign(...,\ constant=True)} and can be used to link
\texttt{Variable}s initial conditions, while trying to use a
\texttt{Parameter} instead is flagged as type errors (red underlining).
The IDE automatically recognizes \texttt{e} as an \texttt{Equation}, and
provides autocompletion of the variables. A tooltip is shown when
composing models, which show the expected variables and their default
values. The IDE highlights wrong names (\texttt{z} is not a name in
\texttt{Model}) and mismatched types (\texttt{x} is \texttt{Variable}
and \texttt{a} must be a number or a \texttt{Constant})}

\end{figure}

@@

\begin{figure}

\begin{minipage}[t]{\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{import}\NormalTok{ numpy }\ImportTok{as}\NormalTok{ np}
\ImportTok{from}\NormalTok{ poincare }\ImportTok{import}\NormalTok{ Simulator}

\NormalTok{sim }\OperatorTok{=}\NormalTok{ Simulator(Oscillator)}
\NormalTok{df }\OperatorTok{=}\NormalTok{ sim.solve(save\_at}\OperatorTok{=}\NormalTok{np.linspace(}\DecValTok{0}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{101}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
\newline
\begin{minipage}[t]{0.50\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df.head()}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[]{@{}lll@{}}
\toprule\noalign{}
& x & v \\
time & & \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
0.00 & 1.000000 & 0.000000 \\
0.05 & 0.998750 & -0.049979 \\
0.10 & 0.995005 & -0.099835 \\
0.15 & 0.988772 & -0.149440 \\
0.20 & 0.980069 & -0.198672 \\
\end{longtable}

}

\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df.plot()}\OperatorTok{;}
\end{Highlighting}
\end{Shaded}

\begin{figure}[H]

{\centering \includegraphics{article_files/figure-latex/cell-9-output-1.pdf}

}

\end{figure}

}

\end{minipage}%

\caption{\label{fig-sim}Simulation of the \texttt{Oscillator} system
from Figure~\ref{fig-second-order}. The output is a
\texttt{pandas.DataFrame} with a column for each variable and the time
as index. It is inspected and plotted with the \texttt{pandas} methods.}

\end{figure}

@@

\begin{figure}

\begin{minipage}[t]{\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{from}\NormalTok{ simbio }\ImportTok{import}\NormalTok{ Compartment, MassAction, Species, RateLaw, initial}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}

{\centering

\begin{minipage}[c]{0.50\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class}\NormalTok{ Model(Compartment):}
    \CommentTok{"""2A {-}\textgreater{} B"""}

\NormalTok{    A: Species }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
\NormalTok{    B: Species }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{0}\NormalTok{)}
\NormalTok{    r }\OperatorTok{=}\NormalTok{ RateLaw(}
\NormalTok{        reactants}\OperatorTok{=}\NormalTok{[}\DecValTok{2} \OperatorTok{*}\NormalTok{ A],}
\NormalTok{        products}\OperatorTok{=}\NormalTok{[B],}
\NormalTok{        rate\_law}\OperatorTok{=}\DecValTok{1}\NormalTok{,}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
%
\begin{minipage}[c]{0.50\linewidth}

{\centering

\[
\begin{cases}
    \frac{dA}{dt} = -2 \\
    \frac{dB}{dt} = +1
\end{cases}
\begin{cases}
    A(0) &= 1 \\
    B(0) &= 0
\end{cases}
\]

}

\end{minipage}%

}

\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}

{\centering

\begin{minipage}[c]{0.50\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class}\NormalTok{ Model(Compartment):}
    \CommentTok{"""2A {-}\textgreater{} B"""}

\NormalTok{    A: Species }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
\NormalTok{    B: Species }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{0}\NormalTok{)}
\NormalTok{    r }\OperatorTok{=}\NormalTok{ MassAction(}
\NormalTok{        reactants}\OperatorTok{=}\NormalTok{[}\DecValTok{2} \OperatorTok{*}\NormalTok{ A],}
\NormalTok{        products}\OperatorTok{=}\NormalTok{[B],}
\NormalTok{        rate}\OperatorTok{=}\DecValTok{1}\NormalTok{,}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
%
\begin{minipage}[c]{0.50\linewidth}

{\centering

\[
\begin{cases}
    \frac{dA}{dt} = -2 A^2 \\
    \frac{dB}{dt} = +1 A^2
\end{cases}
\begin{cases}
    A(0) &= 1 \\
    B(0) &= 0
\end{cases}
\]

}

\end{minipage}%

}

\end{minipage}%

\caption{\label{fig-simbio}A reaction system for species \(A\) and \(B\)
with initial conditions \(1\) and \(0\), respectively. A single reaction
transforming \(2A\) into \(B\) is saved in variable \texttt{r}. The rate
\(1\) is specified directly for \texttt{RateLaw}, and is proportional to
the reactants for \texttt{MassAction}.}

\end{figure}

@@

\begin{figure}

\begin{minipage}[t]{\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{from}\NormalTok{ simbio.io }\ImportTok{import}\NormalTok{ biomodels, sbml}

\NormalTok{model }\OperatorTok{=}\NormalTok{ sbml.load(}\StringTok{"repressilator.sbml"}\NormalTok{)  }\CommentTok{\# from existing SBML file}
\NormalTok{model }\OperatorTok{=}\NormalTok{ biomodels.load\_model(}\StringTok{"BIOMD12"}\NormalTok{)  }\CommentTok{\# from BioModels}
\NormalTok{model}
\end{Highlighting}
\end{Shaded}

Elowitz2000 - Repressilator

type

total

names

variables

6

PX, PY, PZ, X, Y, Z

parameters

17

cell, beta, alpha0, alpha, eff, n, KM, tau\_mRNA, tau\_prot, ...

equations

12

Reaction1, Reaction2, Reaction3, Reaction4, Reaction5, ...

}

\end{minipage}%

\caption{\label{fig-simbio-io}Creation of a model from a local SBML file
or one uploaded to BioModels.}

\end{figure}

@@

\begin{figure}

{\centering \includegraphics{src/performance/figures/performance.png}

}

\caption{\label{fig-runtime}Performance of different softwares to solve
models from the curated section of BioModels. (left) Run time for the
model BIOMD3 as a function of the number of output points. (right) Run
time for different models for 300 output points, using the geometric
mean of the different softwares to order them.}

\end{figure}

>>>>

\input{latex/figure1.tex}

@@

\input{latex/figure2.tex}

@@

\input{latex/figure3.tex}

@@

\input{latex/figure4.tex}

@@

\input{latex/figure5.tex}

@@

\input{latex/figure6.tex}
